Top-down and bottom-up cohesiveness\\in microbial community coalescence

### 2021-05-27 draft/draft_v3.tex
draft v3 done, run through Alvaro

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness in microbial community coalescence}
\author[1 *]{Juan Diaz-Colunga}
\author[1 *]{Nanxi Lu}
\author[1,2 *]{Alicia Sanchez-Gorostiaga}
\author[1]{Chang-Yu Chang}
\author[1]{Helen S. Cai}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1 \Letter]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

\begin{equation}
\frac{dR_\alpha}{dt} = 
- \sum_j N_j c_{j\alpha} R_\alpha
+ \sum_j \sum_\beta N_j c_{j\beta} R_\beta
\left[
l_\beta D_{j\alpha\beta} \frac{w_\beta}{w_\alpha}
\right]
\label{eq:crm-r}
\end{equation}
%
These equations can take slightly different forms in certain cases,
e.g. if the primary resource is supplied continuously instead of at the beggining
of each growth cycle \cite{Marsland2019,Marsland2020},
but they are a good approximation for the community dynamics between consecutive
serial dilutions in our setup.
Here, we assembled \textit{in silico} communities by randomly
sampling a set of species from a pool,
then integrating equations \ref{eq:crm-n} and \ref{eq:crm-r},
diluting the final abundances,
replenishing the primary resource,
and repeating the process until generational equilibrium was achieved
(\methodsref{sim}).
Coalescence simulations were carried out following the same logic,
this time seeding the coalesced communities by mixing the invasive and resident
ones instead of sampling from a species pool.

@@

How do the ideas above scale to more complex and diverse communities?

We reason that the choice of species-specific metabolic architectures is necessary to potentially generate
cohesiveness at the community level during coalescence.
If the secretions of all species were identical (or only slightly different),
higher order cross-feeding effects would be very unspecific:
the establishment of new invasive species
--given that they could outcompete resident taxa within their metabolic niches,
i.e. more effectively feed off the same resource or set of resources as them--
would not alter (or only do so moderately)
the metabolic flows through the rest of the community's cross-feeding network.
On the other hand, said network could undergo a profound and further-reaching restructuring
if the invasive species secreted very different sets of metabolites with respect to the resident ones,
potentially disabling existing niches and/or enabling new ones
where more invaders could be co-selected.
For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
the emergence of cohesiveness in the face of coalescence.
A dense metabolic matrix corresponds to a situation where all species secrete a wide
variety of byproducts.
New-coming invasive species that secrete similar byproducts as resident ones
--even if they do so in different relative amounts--
might only induce moderate quantitative changes in the metabolic fluxes.
But if the sets of secretions are qualitatively different, co-selection of species adapted to
each of those sets becomes possible.
These ideas are supported by experimental observations
suggesting that species with a history of coexistence make up
cohesive communities with highly specific cross-feeding configurations
\cite{Rosenzweig1994,Goldford2018,Estrela2020}.

@@

Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic networks that shape the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
This network is in turn regulated by the supplied primary carbon source
in our minimal laboratory conditions.
This idea is supported by the observation that a Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model also predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
These observations,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

Additional work will be necessary to further clarify the relationship between
metabolic feedbacks, community cohesiveness and ecological co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.
On the other hand, coalescence of communities under different settings
(e.g. in spatially structured environments)
might be governed by additional factors.
Understanding them and quantifying their relative contributions in natural communities
remains an open question in ecology.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig5.pdf}
\caption{\textbf{Community hierarchies determine coalescence outcomes.}
\textbf{a.} Blablabla...
\textbf{b.} Blablabla...
\textbf{b.} Blablabla... ($p=0.083$).}
\label{fig5}
\end{figure}

@@

>>>>

% title, authors, affiliations
\title{Top-down and bottom-up cohesiveness\\in microbial community coalescence}
\author[1 *]{Juan Diaz-Colunga}
\author[1 *]{Nanxi Lu}
\author[1,2 *]{Alicia Sanchez-Gorostiaga}
\author[1]{Chang-Yu Chang}
\author[1]{Helen S. Cai}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1 \Letter]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

\begin{equation}
\frac{dR_\alpha}{dt} = 
- \sum_j N_j c_{j\alpha} R_\alpha
+ \sum_j \sum_\beta N_j c_{j\beta} R_\beta
\left[
l_\beta D_{j\alpha\beta} \frac{w_\beta}{w_\alpha}
\right]
\label{eq:crm-r}
\end{equation}
%
These equations can take slightly different forms in certain cases,
e.g. if the primary resource is supplied continuously instead of at the beggining
of each growth cycle \cite{Marsland2019,Marsland2020}.
They represent a good approximation for the community dynamics between consecutive
serial dilutions in our setup.
Here, we assembled \textit{in silico} communities by randomly
sampling a set of species from a pool,
then integrating equations \ref{eq:crm-n} and \ref{eq:crm-r},
diluting the final abundances,
replenishing the primary resource,
and repeating the process until generational equilibrium was achieved
(\methodsref{sim}).
Coalescence simulations were carried out following the same logic,
this time seeding the coalesced communities by mixing the invasive and resident
ones instead of sampling from a species pool.

@@

How do the ideas above scale to more complex and diverse communities?
In natural microbiomes and in our laboratory cultures,
a large number of species can coexist and cross-feed each other,
giving rise to facilitation networks that are far more dense
than the ones in our minimal model.
However, we argue that some conditions
for the emergence of bottom-up ecological co-selection still apply:
\textit{a)} the byproducts of the invasive cohort should be utilized preferentially
by the invasive dominant (and thus said byproducts need to be species-specific
to minimize overlap with the resident cohort secretions),
and \textit{b)} the invasive dominant should be strongly cross-fed by its cohort.

We reason that the choice of species-specific metabolic architectures is necessary
to potentially generate bottom-up cohesiveness at the community level during coalescence.
If the secretions of all species were identical (or only slightly different),
higher order cross-feeding effects would be very unspecific:
the establishment of new invasive species
--given that they could outcompete resident taxa within their metabolic niches,
i.e. more effectively feed off the same resource or set of resources as them--
would not alter (or only do so moderately)
the metabolic flows through the rest of the community's cross-feeding network.
On the other hand, said network could undergo a deeper and further-reaching restructuring
if the invasive species secreted very different sets of metabolites with respect to the resident ones,
potentially disabling existing niches and/or enabling new ones
where more invaders could be co-selected.
For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
the emergence of bottom-up cohesiveness in the face of coalescence.
A dense metabolic matrix corresponds to a situation where all species secrete a wide
variety of byproducts.
New-coming invasive species that secrete similar byproducts as resident ones
(even if they do so in different relative amounts)
might only induce moderate quantitative changes in the metabolic fluxes.
But if the sets of secretions are qualitatively different, co-selection of species adapted to
each of those sets becomes possible.
These ideas are supported by experimental observations
suggesting that species with a history of coexistence make up
cohesive communities with highly specific cross-feeding configurations
\cite{Rosenzweig1994,Goldford2018,Estrela2020}.
In fact, when we use the MicroCRM to run simulations of community coalescence
but we use a dense metabolic matrix or one that is not species-specific
($D_{i\beta\alpha}=D_{j\beta\alpha}$ for all $i$, $j$, $\alpha$ and $\beta$,
see \hyperref[box1]{Box~1})
we observe virtually no instances of bottom-up ecological co-selection
(\figref{figS4}).

To quantify the degree of cross-feeding from the cohorts towards the dominant,
we defined a community hierarchy metric $h$ as

\begin{equation}
h = \frac{\Delta N_{\mathrm{dom}}^{\mathrm{R1}}}{\Delta N_{\mathrm{dom}}}
\label{eq:h}
\end{equation}
%
where $\Delta N_{\mathrm{dom}}$ represents the overall increase in dominant biomass
between consecutive dilutions in a generationally stable community,
and $\Delta N_{\mathrm{dom}}^{\mathrm{R1}}$
represents the increase in said biomass resulting from the metabolism of the primary resource
($\mathrm{R}_1$) only.
If the dominant was just utilizing the primary resource, the community would be very
hierarchical ($h \sim 1$),
whereas if it was growing mostly on the secretions of other taxa,
it would be more distributed ($h \sim 0$).
We quantified the hierarchies of the resident and invasive communities in our MicroCRM,
finding that $h$ follows a bimodal distribution (\figref[a]{fig5}).
We therefore divided our simulations into four groups according to the relation
of hierarchy between the resident and invasive communities (\figref[b]{fig5}).
For each group, we evaluated the recurrence of bottom-up co-selection,
i.e. the fraction of cases where a dominant that could not invade in isolation
was successful when accompanied by its cohort
(green area of \figref[b]{fig3}).
We found that bottom-up ecological co-selection
was significantly more frequent when the invasive community was
non-hierarchical (\figref[c]{fig5}),
in line with what the minimal model anticipated
(\figref[d-e]{fig4}).

@@

Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic networks that shape the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
This network is in turn regulated by the supplied primary carbon source
in our minimal laboratory conditions.
This idea is supported by the observation that a Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
It also suggests that rarer taxa may play a more prominent role in co-selecting
dominant species when the metabolic fluxes across the community are
distributed rather than hierarchical.
These observations,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions of microbes with one another
and with the environment should be generically expected to produce ecological co-selection
during community coalescence.

Additional work will be necessary to further clarify the relationship between
metabolic feedbacks, community cohesiveness and ecological co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.
Accurately characterizing the metabolic architectures of the species in
this type of laboratory cultures through metabolomics techniques
will also be necessary to verify the relationship between community hierarchy
and the direction of ecological co-selection.
Furthermore, community coalescence under different settings
(e.g. in spatially structured environments)
might be governed by additional factors that we have not considered here.
Understanding them and quantifying their relative contributions in natural communities
remains an open question in ecology.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig5_v2.pdf}
\caption{\textbf{Community hierarchy modulates the recurrence of bottom-up co-selection.}
\textbf{a.} Distribution of community hierarchies for our \textit{in silico} communities.
\textbf{b.} We divided our coalescence simulations into four groups according to the
hierarchies of the resident ($h_\mathrm{R}$) and invasive ($h_\mathrm{I}$) communities
as indicated by the dashed boxes.
For every group, we calculated the fraction of cases where bottom-up co-selection was observed,
i.e. the invasive dominant was unsuccessful when invading in isolation
but successful when invading with its cohort.
\textbf{c.} Bottom-up co-selection of the invasive dominant during coalescence
is significantly more frequent when the invasive community is non-hierarchical.
Error bars representing 95\% confidence intervals
and p-values were computed by bootstrapping
($p<10^{-4}$ where indicated).}
\label{fig5}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.9,keepaspectratio]{figs/figS4.pdf}
\caption{\textbf{Bottom-up ecological co-selection is not observed
when species have similar metabolic architectures.}
We ran simulations of community coalescence following the same
procedure described in the main text, but this time we
used a dense metabolic matrix
(left panel,
sparsity = 0.05 in the Community Simulator package \cite{Marsland2020})
or a species-unspecific metabolic matrix
(right panel,
$D_{i\beta\alpha}=D_{j\beta\alpha}$ for all $i$, $j$, $\alpha$ and $\beta$,
see \hyperref[box1]{Box~1}).
Virtually no bottom-up co-selection is observed in either case.}
\label{figS4}
\end{figure}
